Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2010-08-17 ms.tex
Finished the first draft of the intro.

%\documentclass[apj]{emulateapj}
\documentclass[12pt,preprint]{aastex}

@@

Radiative transfer is ubiquitous in many astrophysical problems,
encompassing topics such as stellar atmospheres, the interstellar
medium (ISM), star formation, galaxy formation, supernovae (SNe) and
cosmology.  Thus it is a well-studied problem
\citep[e.g.][]{Mathews65, Rybicki, Mihalas84, Yorke86}; however, its
treatment in multi-dimensional calculations is difficult because of
the dependence on seven variables --- three spatial, two angular,
frequency, and time.  The non-local nature of the thermal and
hydrodynamical response to radiation sources further adds to the
difficulty.  In the general case, one must consider all radiation
sources at every point in calculation.  Fortunately problems can be
simplified by exploiting symmetries and special cases.  The most basic
case was first considered by \citet{Stroemgren39} for a radiation
source ionizing a static uniform neutral medium.  He found that the
delineation between the neutral and ionized medium was found to be
approximately the mean free path of the ionizing radiation.  His
seminar work was expanded upon by \citet{Spitzer48, Spitzer49,
  Spitzer54} and \citet{Spitzer50}, who showed that the ionizing
radiation heated the medium to $T \sim 10^4$ K.  If the density is
equal on both sides of the ionization front, then this
over-pressurized region would expand and drive a shock outwards
\citep[e.g.][]{Oort54, Schatzman55}.  These early works provided the
basis for the modern topic of radiation hydrodynamics.  A decade
later, radiation hydrodynamical numerical models of \ion{H}{2} regions
in spherical symmetry were developed that described the expansion of
the ionization front and the evolution of its associated shock wave
that carries most of the gas outwards \citep[e.g.][]{Mathews65,
  Lasker66, Hjellming66}.  At the same time, theoretical models of
ionization fronts matured and were classified by \citet{Kahn54} and
\citet{Axford61} as either R-type (rare) or D-type (dense).  R-type
fronts travel supersonically with respect to the neutral gas, whereas
D-type front are subsonic.  Furthermore ``weak'' and ``strong'' R-type
fronts move subsonically and supersonically with respect to the
ionized gas.  The same terminology conversely applies to D-type
fronts.

There are several methods to solve the radiative transfer.

@@

\enzo~is a parallel block-structured AMR \citep{BergerAMR} code that
is publicly available\footnote{http://lca.ucsd.edu/enzo}
\citep{BryanNorman1997, OShea2004}.  In this section, we describe our
parallel implementation of the adaptive ray tracing method into \enzo.
First we explain the programming design of handling the ``photon
packages'' that are traced along the adaptive rays.  We use the terms
photon packages and rays interchangably.  Next we focus on the details
of the radiation hydrodynamics and then the importance of correct
timestepping.  Last we give our parallelization strategy of tracing
rays through an AMR hierarchy.  This implementation is included in the
\texttt{v2.0} public version of \enzo.

@@

In our implementation, photon packages are mono-chromatic and are
assigned a photon type that corresponds whether it is a photon that
(1) ionizes hydrogen, (2) singly ionizes helium, (3) doubly ionizes
helium, (4) has an X-ray energy, or (5) dissociates molecular hydrogen
(Lyman-Werner radiation).  One disadvantage of mono-chromatic rays is
the number of rays increase with the number of frequency bins.
However this allows for early termination of rays that are fully
absorbed, which are likely to have high absorption cross-sections
(e.g. \ion{H}{1} ionizations near 13.6 eV) or a low initial intensity
(e.g. \ion{He}{2} ionizing photons in typical stellar populations).
The other approach used by some groups \citep[e.g.][]{Trac07} is to
store all energy groups in a single ray.  This reduces the number of
the rays generated and the computation associated with the ray
tracing.  Unless the ray dynamically adjusts its memory allocation for
the energy groups as they become depleted, this method is also memory
intensive in the situation where most of the energy groups are
completely absorbed but a few groups still have significant flux.

In practice, we have found that one energy group per photon type is
sufficent to match expected analytical tests.  For example when
modeling Population III stellar radiation \citep[e.g.][for hydrogen
  ionizing radiation only]{Abel07, Wise08b}, we have 3 energy
groups---\ion{H}{1}, \ion{He}{1}, \ion{He}{2}---each with an energy
that equals the average photon energy above the ionization threshold.

@@

In our implementation, a constant timestep is necessary when solving
the time-dependent radiative transfer equation.  It should be small
enough to evolve ionization fronts accurately, as discussed earlier.
The timestep can be based on physical arguments, for example, the
sound-crossing time of an ionized region at $T > 10^4$ K.  To be
conservative, one may choose the sound-crossing time of a cell
\citep[e.g.][]{Abel07, Wise08b}.  Alternatively, the diameter of the
smallest relevant system (e.g., an accretion radius, transition radius
to a D-type ionization front, etc.) in the simulation may be chosen to
calculate the sound-crossing time.

@@

Tests plays an important role in creating and maintaining
computational tools.  In this section, we present tests drawn from the
Cosmological Radiative Transfer Codes Comparison Project
\citep[][hereafter RT06]{RT06}, where results from 11 different
radiative transfer codes compared results in four test problems.  The
codes use various methods for radiation transport: ray tracing with
short, long, and hybrid characteristics, Monte Carlo casting;
ionization front tracking \citep{Alvarez06_IFT}; variable Eddington
Tensor formalism \citep{Gnedin01_OTVET}.  They conducted tests that
investigated (1) the growth of a single Str\"{o}mgren sphere enforcing
isothermal conditions, (2) the same test with an evolving temperature
field, (3) shadowing created by a dense, opticlally thick clump, and
(4) multiple \ion{H}{2} regions in a cosmological density field.  In
all of the tests presented here, we use the method of restricted
neutral fraction changes (\S\ref{sec:dt_hi}) for choosing a radiative
transfer timestep.  We cast 48 rays (HEALPix level 1) from the point
source and require a sampling of at least $\Phi_c = 5.1$ rays per
cell.

>>>>

%\documentclass[apj,onecolumn]{emulateapj}
\documentclass[12pt,preprint]{aastex}

@@

Radiative transfer is ubiquitous in many astrophysical problems,
encompassing topics such as stellar atmospheres, the interstellar
medium (ISM), star formation, galaxy formation, supernovae (SNe) and
cosmology.  Thus it is a well-studied problem
\citep[e.g.][]{Mathews65, Rybicki, Mihalas84, Yorke86}; however, its
treatment in multi-dimensional calculations is difficult because of
the dependence on seven variables --- three spatial, two angular,
frequency, and time.  The non-local nature of the thermal and
hydrodynamical response to radiation sources further adds to the
difficulty.  In the general case, one must consider all radiation
sources at every point in calculation.

Fortunately problems can be simplified by exploiting symmetries and
special cases.  The most basic case was first considered by
\citet{Stroemgren39} for a radiation source ionizing a static uniform
neutral medium.  He found that the delineation between the neutral and
ionized medium was found to be approximately the mean free path of the
ionizing radiation.  His seminar work was expanded upon by
\citet{Spitzer48, Spitzer49, Spitzer54} and \citet{Spitzer50}, who
showed that the ionizing radiation heated the medium to $T \sim 10^4$
K.  If the density is equal on both sides of the ionization front,
then this over-pressurized region would expand and drive a shock
outwards \citep[e.g.][]{Oort54, Schatzman55}.  These early works
provided the basis for the modern topic of radiation hydrodynamics.  A
decade later, the first radiation hydrodynamical numerical models of
\ion{H}{2} regions in spherical symmetry and plane-parallel ionization
fronts were developed \citep[e.g.][]{Mathews65, Lasker66,
  Hjellming66}.  They described the expansion of the ionization front
and the evolution of its associated shock wave that carries most of
the gas away from the source.  At the same time, theoretical models of
ionization fronts matured and were classified by \citet{Kahn54} and
\citet{Axford61} as either R-type (rare) or D-type (dense).  R-type
fronts travel supersonically with respect to the neutral gas, whereas
D-type front are subsonic.  Furthermore ``weak'' and ``strong'' R-type
fronts move subsonically and supersonically with respect to the
ionized gas.  The same terminology conversely applies to D-type
fronts. (add evolution of an \ion{H}{2} region)

In the 1970's and 1980's, algorithmic and computational advances
allowed numerical models to be expanded to two dimensions, mainly
using axi-symmetric to simplify the problem
\citep[e.g.][]{Bodenheimer79, Sandford82, Yorke83}.  One topic that was
studied extensively were champange flows.  Here the source is embedded
in an overdense region, and the \ion{H}{2} region escapes from this
region in one direction.  The interface between the ambient and dense
medium was usually set up to be a constant pressure boundary.  When
the ionization front passes this boundary, the dense, ionized gas is
orders of magnitude out of pressure equilibrium as the temperatures on
both sides of initial boundary are within a factor of a few.  In
response, the gas is accelerated outwards in this direction, creating
a fan-shaped outflow.

Only in the past 15 years, computational resources have become large
enough, along with further algorithmic advances, to cope with the
requirements of three-dimensional calculations.  There are two popular
methods to solve the radiative transfer equation in three-dimensions:
%
\begin{enumerate}
\item Moment methods---The angular moments of the radiation field can
  describe its angular structure, which are related to the energy
  energy, flux, and radiation pressure \citep{Auer70, Norman98}.
  These have been implemented in conjunction with short
  characteristics \citep{Stone92_RT}, with long characteristics
  \citep{Finlator08}, and with a variable Eddington tensor in the
  optically thin limit \citep[OTVET;][]{Gnedin01_OTVET, Petkova09}.
  Moment methods have the advantage of being fast, being independent
  of the number of radiation point sources.  However, they are
  diffusive and result in incorrect shadows in some situations.
\item Ray tracing---Radiation can be propagated along rays that extend
  through the compuational grid \citep[e.g.][]{Mellema98, Razoumov99,
    Abel99, Ciardi00, Ritzerveld03, Whalen05, Rijkhorst05, Mellema06,
    Alvarez06, Trac07, Krumholz07_ART, Paardekooper10} or particle set
  \citep[e.g.][]{Susa06, Johnson07, Pawlik08, Altay08, Hasegawa09,
    Pawlik10}.  In general, these methods are very accurate but
  computationally expensive because the radiation field must be well
  sampled by the rays with respect to the spatial resolution of the
  grid or particles.
\end{enumerate}
%
Until the mid-2000's the vast majority of the three-dimensional
calculations were performed with static density distributions.  One
example is calculating cosmological reionization by post-processing of
density fields from N-body simulations \citep{Ciardi01, Sokasian01,
  McQuinn05, Iliev07}.  Any hydrodynamical response to the radiation
field was thus ignored.  Several radiative transfer codes were
compared in four purely radiative transfer tests in \citet[hereafter
RT06]{RT06}.  Only recently has the radiative transfer equation been
coupled to the hydrodynamics in three-dimensions.  In the second
comparision paper \citep[hereafter RT09]{Iliev09}, results from these
radiation hydrodynamics codes were compared.  Even more rare are ones
that couple it with magneto-hydrodynamics
\citep[e.g.][]{Krumholz07_ART}.  The tests in RT06 and RT09 were kept
relatively simple to ease the comparison.

In this paper, we present our implementation, Enzo+Moray, of adaptive
ray tracing \citep{Abel02_RT} in the cosmological hydrodynamics
adaptive mesh refinement (AMR) code, \enzo~\citep{BryanNorman1997,
  OShea2004}.  The radiation field is coupled to the hydrodynamics
solver at small timescales, enabling it to study radiation
hydrodynamical problems.  We have used this code to investigate the
growth of an \ion{H}{2} region from a 100\Ms~Population III (Pop III)
star \citep{Abel07}, radiative feedback on the formation of high
redshift dwarf galaxies \citep{Wise08_Gal}, the early stages of
reionization from Pop III stars \citep{Wise08_Reion}, ultraviolet
radiation escape fractions from dwarf galaxies before reionization
\citep{Wise09}, radiative negative feedback from accreting Pop III
seed black holes \citep{Alvarez09}, and radiative feedback in
accreting supermassive black holes \citep{Kim11}.  We have included
Enzo+Moray in the latest public release of
\enzo\footnote{http://lca.ucsd.edu/enzo}, and it is also coupled with
the newly added MHD solver in \enzo~\citep{Wang09}.

We have structured this paper as follows.  In Section 2, we describe
the connections between adaptive ray tracing and the radiative
transfer equation.  Furthermore, we detail how physics other than
photo-ionization and photo-heated are included.  We then derive a
geometric correction factor to any ray tracing method to improve
accuracy.  We end the section by describing a new computational
technique to approximate an optically-thin radiation field with ray
tracing and multiple sources.  In Section 3, we cover the details of
our radiation hydrodynamics implementation in \enzo, specifically (1)
the ray tracing algorithms, (2) coupling with the hydrodynamics
solver, (3) several methods to calculate the radiative transfer
timestep, and (4) our parallelization strategy.  We present our
results from the RT06 radiative transfer tests in Section 4.
Afterwards in Section 5, we show the results from the RT09 radiation
hydrodynamics tests.  We expand on these tests to include more
dynamical and complex setups to demonstrate the flexibility and high
fidelity of Enzo+Moray.  Section 6 reports the results from spatial,
angular, frequency, and temporal resolution tests.  In Section 7, we
illustrate the improvements from the geometric correction factor and
our optically-thin approximation.  We also show the effects of X-ray
radiation and radiation pressure in this section.  Finally in Section
8, we demonstrate the parallel scalability of Enzo+Moray.  Last
Section 9 summarizes our method and results.

@@

\enzo~is a parallel block-structured AMR \citep{BergerAMR} code that
is publicly available \citep{BryanNorman1997, OShea2004}.  In this
section, we describe our parallel implementation of the adaptive ray
tracing method into \enzo.  First we explain the programming design of
handling the ``photon packages'' that are traced along the adaptive
rays.  We use the terms photon packages and rays interchangably.  Next
we focus on the details of the radiation hydrodynamics and then the
importance of correct timestepping.  Last we give our parallelization
strategy of tracing rays through an AMR hierarchy.  This
implementation is included in the \texttt{v2.0} public version of
\enzo.

@@

In our implementation, photon packages are mono-chromatic, i.e. energy
groups \citep[][Ch. 6]{Mihalas84}, and are assigned a photon type that
corresponds whether it is a photon that (1) ionizes hydrogen, (2)
singly ionizes helium, (3) doubly ionizes helium, (4) has an X-ray
energy, or (5) dissociates molecular hydrogen (Lyman-Werner
radiation).  One disadvantage of mono-chromatic rays is the number of
rays increase with the number of frequency bins.  However this allows
for early termination of rays that are fully absorbed, which are
likely to have high absorption cross-sections (e.g. \ion{H}{1}
ionizations near 13.6 eV) or a low initial intensity (e.g. \ion{He}{2}
ionizing photons in typical stellar populations).  The other approach
used by some groups \citep[e.g.][]{Trac07} is to store all energy
groups in a single ray.  This reduces the number of the rays generated
and the computation associated with the ray tracing.  Unless the ray
dynamically adjusts its memory allocation for the energy groups as
they become depleted, this method is also memory intensive in the
situation where most of the energy groups are completely absorbed but
a few groups still have significant flux.

In practice, we have found that one energy group per photon type is
sufficent to match expected analytical tests.  For example when
modeling Population III stellar radiation \citep[e.g.][for hydrogen
  ionizing radiation only]{Abel07, Wise08_Gal}, we have 3 energy
groups---\ion{H}{1}, \ion{He}{1}, \ion{He}{2}---each with an energy
that equals the average photon energy above the ionization threshold.

@@

In our implementation, a constant timestep is necessary when solving
the time-dependent radiative transfer equation.  It should be small
enough to evolve ionization fronts accurately, as discussed earlier.
The timestep can be based on physical arguments, for example, the
sound-crossing time of an ionized region at $T > 10^4$ K.  To be
conservative, one may choose the sound-crossing time of a cell
\citep[e.g.][]{Abel07, Wise08_Gal}.  Alternatively, the diameter of the
smallest relevant system (e.g., an accretion radius, transition radius
to a D-type ionization front, etc.) in the simulation may be chosen to
calculate the sound-crossing time.

@@

Tests plays an important role in creating and maintaining
computational tools.  In this section, we present tests drawn from the
Cosmological Radiative Transfer Codes Comparison Project \citep{RT06},
where results from 11 different radiative transfer codes compared
results in four test problems.  The codes use various methods for
radiation transport: ray tracing with short, long, and hybrid
characteristics, Monte Carlo casting; ionization front tracking
\citep{Alvarez06_IFT}; variable Eddington Tensor formalism
\citep{Gnedin01_OTVET}.  They conducted tests that investigated (1)
the growth of a single Str\"{o}mgren sphere enforcing isothermal
conditions, (2) the same test with an evolving temperature field, (3)
shadowing created by a dense, opticlally thick clump, and (4) multiple
\ion{H}{2} regions in a cosmological density field.  In all of the
tests presented here, we use the method of restricted neutral fraction
changes (\S\ref{sec:dt_hi}) for choosing a radiative transfer
timestep.  We cast 48 rays (HEALPix level 1) from the point source and
require a sampling of at least $\Phi_c = 5.1$ rays per cell.
